Equations & R Code

### 2016-12-13 manuscript/coexistence-stability_ms.tex
updates equation 2; maybe wrong
first edits from FRA added

However, the ability of biodiversity-ecosystem functioning (BEF)
experiments to accurately represent real-world dynamics is debated
(Wardle 2016, Eisenhauer et al. 2016). Much of the debate centers around
the fact BEF experimental protocols do not allow species gains to offset
species losses. Theoretical work on diversity-stability relationships
typically suffers from the same limitation: it recognizes the role of
environmental variability in driving population fluctuations which
destabilize ecosystems, but ignores the potential for environmental
variability to promote species richness and thereby help stabilize
ecosystems (Loreau 2010, but see Chesson et al. 2001).

@@

We developed a semi-discrete consumer-resource model that allows many
species to coexist on one resource by either the storage effect or
relative nonlinearity. In our model, the consumer can be in one of
two-states: a dormant state \(D\) and a live state \(N\). The dormant
state could represent, for example, the seedbank of an annual plant.
Transitions between \(N\) and \(D\) occur at discrete intervals \(\tau\)
with continuous-time consumer-resource dynamics between discrete
transitions. Thus, our model is formulated as ``pulsed differential
equations'' (Pachepsky et al. 2008, Mailleret and Lemesle 2009, Mordecai
et al. 2016). For clarity we refer to \(\tau\) as years and the growing
time between years as seasons with daily (\(t\)) time steps.

@@

\noindent where \(D(\tau_k)\), \(N(\tau_k)\), and \(R(\tau_k)\) are the
abundances of each state at the end of growing season \emph{k} and
\(\tau_k^+\) denotes the beginning of growing season \(k=1\). The
activation of dormant biomass to live biomass is controlled by
\(\gamma\), which is year (\emph{k}) and species (\emph{i}) specific.
Dormant biomass is equal to a constant fraction (\(\alpha\)) of live
biomass at the end of the previous season (\(N_{i}(\tau_k)\)), plus
survival (\(1-\eta_i\)) of dormant biomass (\(D_{i}(\tau_k)\)) at the
end of the previous year and dormant biomass remaining after live
biomass activation (\(D_{i}(\tau_k)(1-\gamma_{i,\tau_k})\)). Live
biomass is equal to newly activated dormant biomass
(\(\gamma_{i,t}[D_{i}(\tau_k)\)), minus some fraction of live biomass
that is converted to dormant biomass (\((1-\alpha_i)N_{i}(\tau_k)\)) We
assume the resource pool is not replenished within a growing season.
Resource replenishment occurs between growing seasons, and the resource
pool (\emph{R}) at the start of the growing season \emph{k}+1 is
\(R(\tau_k^+) = R^+\), where \(R^+\) is a random resource pulse drawn
from a log-normal distribution with mean \(\mu(R^+)\) and standard
deviation \(\sigma(R^+)\). Model parameters and notation are described
in Table 1.

@@

\begin{align}
\Sigma(\gamma) = 
\begin{bmatrix}
\sigma^2_{E} & \rho_{1,2}\sigma^2_{E} & \rho_{1,3}\sigma^2_{E} & \rho_{1,4}\sigma^2_{E} \\
\rho_{2,1}\sigma^2_{E} & \sigma^2_{E} & \rho_{2,3}\sigma^2_{E} & \rho_{2,4}\sigma^2_{E} \\
\rho_{3,1}\sigma^2_{E} & \rho_{3,2}\sigma^2_{E} & \sigma^2_{E}  & \rho_{3,4}\sigma^2_{E} \\
\rho_{4,1}\sigma^2_{E} & \rho_{4,2}\sigma^2_{E} & \rho_{4,3}\sigma^2_{E} & \sigma^2_{E}  \\
\end{bmatrix}
\end{align}

\noindent where \(\sigma^2_{E}\) is the variance of the environmental
cue and \(\rho_{i,j}\) is the correlation between between the species
\emph{i}'s and species \emph{j}'s transition rates. \(\rho\) must be
less than 1 for stable coexistence, and in all simulations we placed
that constraint that \(\rho_{i,j} = \rho_{j,i}\) for each species pair.
The inferior competitor has the strongest potential to persist when
\(\rho=-1\) (perfectly uncorrelated transition rates). We used the
\texttt{R} function \texttt{mvrnorm} to generate sequences of
(un)correlated variates \emph{E} that we converted to germination rates
in the 0-1 range: \(\gamma = e^E / 1 + e^E\). Note that
\(\Sigma(\gamma)\) must be positive definite. So, after defining
\(\Sigma(\gamma)\) with all \(\rho_{i,j}\)s and \(\sigma^2_{E}\), we
used the \texttt{nearPD} function from the \texttt{Matrix} package in
\texttt{R} to coerce the variance-covariance matrix to be positive
definite.

@@

For example, in Fig. 1C we show two species' resource uptake curves that
are relatively nonlinear. Species B has the lowest \(R^*\) and would
competitively exclude species A in the absence of environmental
fluctuations. But, fluctuating resource supplies can benefit species A
because it can take advantage of relatively high resource levels due its
higher saturation point. Stable coexistence is only possible, however,
if each species creates a disadvantage for itself when abundant. This
occurs in our model because when a resource conservative species (e.g.,
species B in Fig. 1C) is abundant, resource levels will remain high for
a longer period of time because its draw down of resources starates.
Likwise, when a resource acquisitive species (e.g., species A in Fig.
1C) is abundant, it quickly draws down resources to levels that favor
resource conservative species. Such reciprocity ensures species can
increase when rare and stablizes coexistence (Armstrong and McGehee
1980, Chesson 2000a, Chesson et al. 2004).

@@

Under relative nonlinearity, species' resource response curves (Fig.
S1-1) reflect traits that determine the intrinsic stability of each
species. Therefore, we ran two sets of simulations for relative
nonlinearity: one where the species pools increased from stable to
unstable species and vice versa. For example, if species A is the most
stable species and species D is the least stable we ran simulations
where the species pool increased from one to four species as A then B
then C then D. We then ran simulations with that order reversed.

All simulations were run for 5,000 seasons with 20-day growing seasons.
We averaged biomass over the growing season, and those yearly values
were used to calculate total community biomass in each year. After
discarding an initial 500 seasons to reduce transient effects on our
results, we calculated the coffecient of variation (\emph{CV}) of summed
species biomass through time, which represents ecosystem variability,
the inverse of ecosystem stability. We calculated species richness as
the number of species whose average biomass was greater than 1 over the
course of the simulation. Parameter values for specific results are
given in figure captions. Within-season dynamics were solved given
initial conditions using the package \texttt{deSolve} (Soetaert et al.
2010) in \texttt{R} (Team 2013). \texttt{R} code for our model function
is in the Supporting Information Appendix S1. All model code has been
deposited on Figshare (\emph{link}) and is available on GitHub at
\url{http://github.com/atredennick/Coexistence-Stability}.

@@

The dampening effect of fluctuation-dependent coexistence on increasing
environmental variability depends on the specific traits (parameter
values) of the species in the regional pool. Asymmetric competition
makes it more difficult for new species to enter the local community
(Fig. 3; compare top and bottom panels). Relatively high competition
also increases the rate at which ecosystem \emph{CV} increases with
environmental variance (Fig. S1-3) because the abundance of inferior
competitors is reduced and therefore does not influence ecosystem
\emph{CV} as much as when competition is symmetric. The correlation of
species' environmental responses also mediates the relationship between
environmental variance, species richness, and ecosystem \emph{CV}: lower
correlations make it easier for new species to enter the community and
unique environmental responses are always stabilizing (Fig. 3).

@@

Theory developed for biodiversity-ecosystem function experiments
emphasizes that increases in species richness should reduce ecosystem
variability. Consistent with theoretical expectations from models in
which species coexistence is maintained by fluctuation-independent
mechanisms, we produced the typical negative diversity-variability
relationship (Fig. 2B,D) from a model with fluctation-dependent species
coexistence (also see Chesson et al. 2001). Likewise, our results from
the species removal simulations are consistent with results from
biodiversity-ecosystem functioning experiments showing a negative
relationship between species richness and ecosystem variability. This is
encouraging because species almost certainly coexist by some combination
of fluctuation-independent (e.g., resource partitioning) and
fluctuation-dependent mechanisms. By extending theory to communities
where species richness is explicitly maintained by temporal variability,
we have gained confidence that experimental findings are generalizable
to many communities. In other words, in local settings where
environmental variability is relatively homogenous, reductions in the
number of species will reduce the stability of ecosystem functioning,
regardless of how coexistence is maintained.

When we allowed a gradient of environmental variability to determine
species coexistence, we discovered a positive relationship between
species richness and ecosystem variability (Fig. 2A,C). While surprising
when viewed through the lens of previous theory and experimental
findings, such a relationship is a direct consequence of how diversity
can be maintained in fluctuating environments. The storage effect and
relative nonlinearity both require envrionmental fluctuations to allow
niche differentiation between species pairs (Chesson 2000a). Therefore,
species coexistence gains strength, for both mechanisms, as the
environment becomes more variable (Fig. S1-2), but the increase in
environmental variability also increases ecosystem variability.

Our results may explain why deviations from the negative
diversity-variability relationship often come from observational studies
(Jiang and Pu 2009). Observational studies must rely on natural
diversity gradients, and if species richness depends on environmental
variability, it is entirely possible to observe positive
diversity-variability relationships. For example, Sasaki and Lauenroth
(2011) found a negative relationship between species richness and the
temporal stability of plant abundance (a positive diversity-variability
relationship) in a semi-arid grassland. Their data came from a six sites
that were 6 km apart. While Sasaki and Lauenroth explained their results
in terms of dominant species' effects, it is also possible that each
site experienced slightly different levels of environmental variability
that influenced species coexistence. DeClerck et al. (2006) also found a
positive diversity-variability when sampling conifer richness and the
variability of productivity across a large spatial gradient in the
Sierra Nevada.

>>>>

However, the ability of biodiversity-ecosystem functioning (BEF)
experiments to accurately represent real-world dynamics is debated
(Wardle 2016, Eisenhauer et al. 2016). Much of the debate centers around
the fact that BEF experimental protocols do not allow species gains to
offset species losses. Theoretical work on diversity-stability
relationships typically suffers from the same limitation: it recognizes
the role of environmental variability in driving population fluctuations
which destabilize ecosystems, but ignores the potential for
environmental variability to promote species richness and thereby help
stabilize ecosystems (Loreau 2010, but see Chesson et al. 2001).

@@

We developed a semi-discrete consumer-resource model that allows many
species to coexist on one resource by either the storage effect or
relative nonlinearity. In our model, the consumer can be in one of
two-states: a dormant state \(D\) and a live state \(N\). The dormant
state could represent, for example, the seed bank of an annual plant.
Transitions between \(N\) and \(D\) occur at discrete intervals \(\tau\)
with continuous-time consumer-resource dynamics between discrete
transitions. Thus, our model is formulated as ``pulsed differential
equations'' (Pachepsky et al. 2008, Mailleret and Lemesle 2009, Mordecai
et al. 2016). For clarity we refer to \(\tau\) as years and the growing
time between years as seasons with daily (\(t\)) time steps.

@@

\noindent where \(D(\tau_k)\), \(N(\tau_k)\), and \(R(\tau_k)\) are the
abundances of each state at the end of growing season \emph{k} and
\(\tau_k^+\) denotes the beginning of growing season \(k\). The
activation of dormant biomass to live biomass is controlled by
\(\gamma\), which is year (\emph{k}) and species (\emph{i}) specific.
Dormant biomass is equal to a constant fraction (\(\alpha\)) of live
biomass at the end of the previous season (\(N_{i}(\tau_k)\)), plus
survival (\(1-\eta_i\)) of dormant biomass (\(D_{i}(\tau_k)\)) at the
end of the previous year and dormant biomass remaining after live
biomass activation (\(D_{i}(\tau_k)(1-\gamma_{i,\tau_k})\)). Live
biomass is equal to newly activated dormant biomass
(\(\gamma_{i,t}D_{i}(\tau_k)\)), minus some fraction of live biomass
that is converted to dormant biomass (\((1-\alpha_i)N_{i}(\tau_k)\)) We
assume the resource pool is not replenished within a growing season.
Resource replenishment occurs between growing seasons, and the resource
pool (\emph{R}) at the start of the growing season \emph{k}+1 is
\(R(\tau_k^+) = R^+\), where \(R^+\) is a random resource pulse drawn
from a log-normal distribution with mean \(\mu(R^+)\) and standard
deviation \(\sigma(R^+)\). Model parameters and notation are described
in Table 1.

@@

\begin{align}
\Sigma(\gamma) = 
\begin{bmatrix}
1 & \rho_{1,2} & \rho_{1,3} & \rho_{1,4} \\
\rho_{2,1} & 1 & \rho_{2,3} & \rho_{2,4} \\
\rho_{3,1} & \rho_{3,2} & 1  & \rho_{3,4} \\
\rho_{4,1} & \rho_{4,2} & \rho_{4,3} & 1  \\
\end{bmatrix}
\sigma_{E}^2
\end{align}

\noindent where \(\sigma^2_{E}\) is the variance of the environmental
cue and \(\rho_{i,j}\) is the correlation between species \emph{i}'s and
species \emph{j}'s transition rates. \(\rho\) must be less than 1 for
stable coexistence, and in all simulations we placed that constraint
that all \(\rho_{i,j}\)'s are equal. In a two-species community, the
inferior competitor has the strongest potential to persist when
\(\rho=-1\) (perfectly uncorrelated transition rates). However, in a
four-species community the minimum possible correlation among species is
-1/3 given our constraints that all \(\rho\)'s are equal and that
\(\Sigma(\gamma)\) must be positive-definite. We used the \texttt{R}
function \texttt{mvrnorm} to generate sequences of (un)correlated
variates \emph{E} that we converted to germination rates in the 0-1
range: \(\gamma = e^E / \left(1 + e^E \right)\).

@@

For example, in Fig. 1C we show two species' resource uptake curves that
are relatively nonlinear. Species B has the lowest \(R^*\) and would
competitively exclude species A in the absence of environmental
fluctuations. But, fluctuating resource supplies can benefit species A
because it can take advantage of relatively high resource levels due its
higher saturation point. Stable coexistence is only possible, however,
if each species creates a disadvantage for itself when abundant. This
occurs in our model because when a resource conservative species (e.g.,
species B in Fig. 1C) is abundant, resource levels will remain high for
a longer period of time because its draw down of resources saturates.
Likewise, when a resource acquisitive species (e.g., species A in Fig.
1C) is abundant, it quickly draws down resources to levels that favor
resource conservative species. Such reciprocity ensures species can
increase when rare and stabilizes coexistence (Armstrong and McGehee
1980, Chesson 2000a, Chesson et al. 2004).

@@

Under relative nonlinearity, species' resource response curves (Fig.
S1-1) reflect traits that determine the variability each species' growth
through time. ``Stable'' species have conservative resource uptake
curves, whereas ``unstable'' species have highly nonlinear resource
uptake curves with a large difference in growth between low and high
resource availability. Therefore, we ran two sets of simulations for
relative nonlinearity: one where the species pools increased from stable
to unstable species and vice versa. For example, if species A is the
most stable species and species D is the least stable we ran simulations
where the species pool increased from one to four species as A then B
then C then D. We then ran simulations with that order reversed.

All simulations were run for 5,000 seasons with 100-day growing seasons.
We averaged biomass over the growing season, and those yearly values
were used to calculate total community biomass in each year. After
discarding an initial 500 seasons to reduce transient effects on our
results, we calculated the coefficient of variation (\emph{CV}) of
summed species biomass through time, which represents ecosystem
variability, the inverse of ecosystem stability. We calculated species
richness as the number of species whose average biomass was greater than
1 over the course of the simulation. Parameter values for specific
results are given in figure captions. Within-season dynamics were solved
given initial conditions using the package \texttt{deSolve} (Soetaert et
al. 2010) in \texttt{R} (Team 2013). \texttt{R} code for our model
function is in the Supporting Information Appendix S1. All model code
has been deposited on Figshare (\emph{link}) and is available on GitHub
at \url{http://github.com/atredennick/Coexistence-Stability}.

@@

The dampening effect of fluctuation-dependent coexistence on increasing
environmental variability depends on the specific traits (parameter
values) of the species in the regional pool. Asymmetric competition
makes it more difficult for new species to enter the local community,
but once they do enter, variability is similar (Fig. 3; compare top and
bottom panels). Relatively high competition also increases the rate at
which ecosystem \emph{CV} increases with environmental variance (Fig.
S1-3). This is because the abundance of inferior competitors is reduced
and do not influence ecosystem \emph{CV} as much as when competition is
symmetric. The correlation of species' environmental responses also
mediates the relationship between environmental variance, species
richness, and ecosystem \emph{CV}: lower correlations make it easier for
new species to enter the community and unique environmental responses
are always stabilizing (Fig. 3).

@@

Theory developed for biodiversity-ecosystem function experiments
emphasizes that increases in species richness should reduce ecosystem
variability. Consistent with theoretical expectations from models in
which species coexistence is maintained by fluctuation-independent
mechanisms, we produced the typical negative diversity-variability
relationship (Fig. 2B,D) from a model with fluctuation-dependent species
coexistence (also see Chesson et al. 2001). Likewise, our results from
the species removal simulations are consistent with results from
biodiversity-ecosystem functioning experiments showing a negative
relationship between species richness and ecosystem variability. This is
encouraging because species almost certainly coexist by some combination
of fluctuation-independent (e.g., resource partitioning) and
fluctuation-dependent mechanisms. By extending theory to communities
where species richness is explicitly maintained by temporal variability,
we have gained confidence that experimental findings are generalizable
to many communities. In other words, in local settings where
environmental variability is relatively homogeneous, reductions in the
number of species will reduce the stability of ecosystem functioning,
regardless of how coexistence is maintained.

When we allowed a gradient of environmental variability to determine
species coexistence, we discovered a positive relationship between
species richness and ecosystem variability (Fig. 2A,C). While surprising
when viewed through the lens of previous theory and experimental
findings, such a relationship is a direct consequence of how diversity
can be maintained in fluctuating environments. The storage effect and
relative nonlinearity both require environmental fluctuations to allow
niche differentiation between species pairs (Chesson 2000a). Therefore,
species coexistence gains strength, for both mechanisms, as the
environment becomes more variable (Fig. S1-2), but the increase in
environmental variability also increases ecosystem variability.

Our results may explain why deviations from the negative
diversity-variability relationship often come from observational studies
(Jiang and Pu 2009). Observational studies must rely on natural
diversity gradients, and if species richness depends on environmental
variability, it is entirely possible to observe positive
diversity-variability relationships. For example, Sasaki and Lauenroth
(2011) found a negative relationship between species richness and the
temporal stability of plant abundance (a positive diversity-variability
relationship) in a semi-arid grassland. Their data came from a six sites
that were 6 km apart. While Sasaki and Lauenroth explained their results
in terms of dominant species' effects, it is also possible that each
site experienced sufficiently different levels of environmental
variability to influence species coexistence. DeClerck et al. (2006)
also found a positive diversity-variability when sampling conifer
richness and the variability of productivity across a large spatial
gradient in the Sierra Nevada.
